### 2023-11-17 paper.tex
add fig

\new{
Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions.
This demonstrates that our VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations.
An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}F).
}

\new{
We repeated this benchmark on an additional \todo{3(?)} published single-cell methylome data sets.
These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:todo}).
}

@@

Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
Overall, our default parameter combination´ provided good results and fast compute times in both data sets.

@@

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
\todo{Rewrite this part to discuss other tools: The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.}
\new{
Our benchmarks demonstrate that selecting informative genomic features is a crucial aspect of scBS data analysis.
Unlike other tools for scBS data analysis \citep{}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
Furthermore, \todo{DMR detection and residuals}
}
In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
Depending on the research question at hand, individual DMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.

@@

\subsection{Analysis of scBS datasets for benchmarks}
To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, \new{stepsize $=100$}, variance threshold $= 0.02$, minimum cell requirement $=6$).
Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
The same analysis was repeated for mouse scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
For each data set, we thus obtained four methylation matrices:
either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
All four matrices were then subjected to PCA.
In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca".
To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
	    \includegraphics[width=0.4\textwidth]{figures/SFig_benchmark.png}
	\end{center}
	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex}.\\
	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
	}
	\label{fig:score_luo}
\end{minipage}
\vspace{6ex}

\noindent%
\begin{minipage}{\linewidth}
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/SFig_parameter_sweep.png}
    \end{center}
    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
    \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
    VMRs were detected with \texttt{scbs scan} using various sliding window bandwidths or variance thresholds, and on sub-samples of the full data sets.
    }
    \label{fig:sweep}
\end{minipage}

>>>>

\new{
Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions.
This demonstrates that our VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations.
An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}).
}

\new{
We repeated this benchmark on an additional \todo{3(?)} published single-cell methylome data sets.
These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
}

@@

Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
For the \new{CpG-methylation} data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
\new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation, we furthermore performed VMR detection with default options on the mCH-portion of this data set.}
In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
Overall, our default parameter combination´ provided good results and fast compute times in both data sets.

@@

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
\todo{Rewrite this part to discuss other tools: The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.}
\new{
Our benchmarks demonstrate that selecting informative genomic features is a crucial aspect of scBS data analysis.
Unlike other tools for scBS data analysis \citep{}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
Furthermore, \todo{other tools don't offer DMR detection and our residuals}.
\todo{scbs is compatible with other tools that are up- or downstream of its capabilities, as we've shown here for MOFA+.}
}
In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
Depending on the research question at hand, individual DMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.

@@

\subsection{Analysis of scBS datasets for benchmarks}
To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
The same analysis was repeated for mouse scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
For each data set, we thus obtained four methylation matrices:
either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
All four matrices were then subjected to PCA.
In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca".
To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
	    \includegraphics[width=0.4\textwidth]{figures/SFig_benchmark.png}
	\end{center}
	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex.}\\
	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
	}
	\label{fig:score_luo}
\end{minipage}
\vspace{6ex}

\noindent%
\begin{minipage}{\linewidth}
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/SFig_parameter_sweep.png}
    \end{center}
    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types.}\\
    \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
    VMRs were detected with \texttt{scbs scan} using various sliding window bandwidths or variance thresholds, and on sub-samples of the full data sets.
    }
    \label{fig:sweep}
\end{minipage}

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
		\includegraphics[width=.8\textwidth]{figures/SFig_benchmark_resources.png}
	\end{center}
	\captionof{figure}{\small \new{\textbf{VMRs are as informative as regulatory regions, but more resource-efficient.}\\
	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions.
	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.
	\textbf{(C-D)} Runtime (top) and RAM usage (bottom) of the two dimensionality reduction techniques shown in A and B.
	}}
	\label{fig:resource_usage}
\end{minipage}
